---
title: "Starting With Weathering Trends"
author: "Aaron Rendahl"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Starting With Weathering Trends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## A basic example fitting a single element pair

```{r}
library(WeatheringTrends)
data(welldata)
(fit1 <- FitElementRatio("Sr", "Zr", "depth.top", welldata))
plot(fit1)
```

```{r}
coef(fit1)
coef(fit1, type="par")
coef(fit1, type="par.long")
```

## Allows for enrichment of elements

Fake this by simply turning around mobile and immobile.
```{r}
(fit2 <- FitElementRatio("Zr", "Sr", "depth.top", welldata))
plot(fit2)
```

## Can fit multiple fits at once, turn off profiling of depth for this

```{r, fig.show='hold', fig.width=8, fig.height=10}
fits <- FitElementRatios(c("Sr", "Pb", "CaO"), c("Zr", "V"), "depth.top", welldata, profile=FALSE, verbose=FALSE)
plot(fits)
```

Can also get coefficients here easily too
```{r}
coef(fits)
coef(fits, type="par")
coef(fits, type="par.long")
```

## Testing this out...

Can also change the fit so it's linear on the log(ratio) scale instead of on the ratio scale.

Original method, linear on the ratio scale.
```{r, fig.width=8, fig.height=4}
(fita <- FitElementRatio("Sr", "Zr", "depth.top", welldata, profile=FALSE))
par(mfrow=c(1,2))
plot(fita)
plot(fita, log=FALSE)
```

Or linear on the log scale.
```{r, fig.width=8, fig.height=4}
(fitb <- FitElementRatio("Sr", "Zr", "depth.top", welldata, loglinear=TRUE, profile=FALSE))
par(mfrow=c(1,2))
plot(fitb)
plot(fitb, log=FALSE)
```

## Computing tau directly

Can either take the average on the ratio scale or on the logratio scale.
Plotting is done directly, so far.

```{r, fig.width=8, fig.height=4}
library(WeatheringTrends)
data(welldata)
foo1 <- FitTau("Sr", "Zr", "depth.top", welldata, cutoff=7.5)
with(foo1$data, {
  plot(tau, -depth, main="tau")
  abline(h=-foo1$cutoff, v=0)
})
